---
title: "Antimicrobial Stewardship: Co-amoxiclav Prescribing in England and C. difficile Risk"
author: "Marco Sorbona, PhD"
date: "April 2026"
format: 
  html:
    toc: true
    code-fold: false
    theme: cosmo
editor: visual
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: load-libraries
#| include: false

# Load libraries
library(tidyverse)
library(lubridate)
library(plotly)
library(knitr)
library(DT)
library(scales)

# Load the clean, analysis-ready data
coamox_data <- readRDS("data/processed/coamox_data_curated.rds")
```

## Introduction

Co-amoxiclav (co-amoxicillin-clavulanate) is a broad-spectrum antibiotic classified as one of the "4C" antibiotics that significantly increase the risk of Clostridioides difficile infection (CDI). This analysis investigates:

1.  Trends in co-amoxiclav prescribing across England (2021–2026)

2.  Regional variation in prescribing

3.  The link to a previously detected C. difficile outbreak signal in the East Midlands

## Data Overview

```{r}
# Data Structure
glimpse(coamox_data)

# Summary statistics
coamox_data |> 
  summarise(
    start_date = min(date),
    end_date = max(date),
    total_organisations = n_distinct(organisation_code),
    total_prescribing_items = sum(prescribing_items, na.rm = TRUE),
    total_cost = sum(prescribing_cost, na.rm = TRUE)
  ) |> 
  kable(caption = "Dataset summary")
```

## National Trends

Co-amoxiclav prescribing in England shows two distinct patterns:

### 1. Prescribing volume has remained stable

The faceted plot (upper panel) shows prescription items fluctuating between **90,000 and 110,000 per month** from February 2021 to January 2026. There is no significant upward or downward trend in prescribing volume.

### 2. Total cost has been volatile

The same plot (lower panel) reveals dramatic cost variation:

-   **2021-2022**: Stable monthly costs (\~£250,000-300,000)
-   **Late 2022 - 2023**: A **sharp spike** beginning November 2022, peaking in December 2022 (£1.06 million) and remaining elevated through 2023
-   **2024-2026**: Gradual decline to lowest levels in January 2026 (£211,000)

```{r}
#| label: national-trends-faceted
#| fig-cap: "Co-amoxiclav prescribing in England (2021-2026)"

# National Monthly Trend
national_trend <- coamox_data |> 
  group_by(date) |> 
  summarise(
    total_prescribing_items = sum(prescribing_items, na.rm = TRUE),
    total_cost = sum(prescribing_cost, na.rm = TRUE),
    .groups = "drop"
  )

# Reshape data for faceting
national_long <- national_trend |> 
  pivot_longer(
    cols = c(total_prescribing_items, total_cost),
    names_to = "metric",
    values_to = "value"
  ) |> 
  mutate(
    metric = case_when(
      metric == "total_prescribing_items" ~ "Prescription items",
      metric == "total_cost" ~ "Total cost (£)"
    )
  )

# Plot with free y-axes
p_faceted <- ggplot(national_long, aes(x = date, y = value)) +
  geom_line(color = "steelblue", linewidth = 0.8) +
  geom_point(size = 0.5) +
  facet_wrap(~metric, scales = "free_y", ncol = 1) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Co-amoxiclav Prescribing in England",
    subtitle = "Monthly prescription items and total cost (Feb 2021 – Jan 2026)",
    x = "Date",
    y = ""
  ) +
  theme_minimal()

ggplotly(p_faceted)
```

### The two-y-axis plot tells the whole story

The combined plot shows the **cost spike visually overwhelming the prescribing volume**: this is not a visualisation flaw but the actual finding. At the peak, costs were **5 times higher** than baseline while prescribing volume remained unchanged.

**Interpretation:** The NHS experienced a dramatic increase in co-amoxiclav costs in 2022-2023, likely due to:

-   Supply chain disruptions

-   Manufacturer price increases

-   Generic market changes

The subsequent decline to record-low costs in 2025-2026 suggests successful procurement interventions or new generic entry.

**Stewardship implication:** Prescribing volume has been stable throughout. Future stewardship should focus on **appropriate use** and **cost-effective prescribing** without reducing necessary treatments.

```{r}
#| label: national-trends
#| fig-cap: "Co-amoxiclav prescribing in England (2021-2026)"

# Plot with two y-axes (items and cost)
ggplot(national_trend, aes(x = date)) +
  geom_line(aes(y = total_prescribing_items, color = "Prescription items"), linewidth = 0.8) +
  geom_line(aes(y = total_cost, color = "Total cost (£k)"), linewidth = 0.8) +
  scale_y_continuous(
    name = "Prescription items",
    labels = scales::comma,
    sec.axis = sec_axis(~ . * 1, name = "Total cost (£)", labels = scales::comma)
  ) +
  scale_color_manual(values = c("Prescription items" = "steelblue", "Total cost (£k)" = "darkred")) +
  labs(
    title = "Co-amoxiclav Prescribing in England",
    subtitle = "Monthly prescription items (Feb 2021 - present)",
    x = "Date",
    y = "Total prescription items",
    color = ""
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    axis.title.y.right = element_text(color = "darkred"),
    axis.text.y.right = element_text(color = "darkred"),
    axis.title.y.left = element_text(color = "steelblue"),
    axis.text.y.left = element_text(color = "steelblue")
  )
```

### Cost per item

```{r}
#| label: cost-per-item
#| fig-cap: "Cost per prescription"

cost_per_item <- national_trend |> 
  mutate(
    cost_per_item = total_cost / total_prescribing_items,
    year = year(date)
  )

p_cost_per_item <- ggplot(cost_per_item,aes(x = date, y = cost_per_item)) +
  geom_line(color = "darkred", linewidth = 0.8) +
  geom_point(size = 1) +
  scale_y_continuous(
    labels = scales::dollar_format(prefix = "£") 
  ) +
  labs(
    title = "Co-amoxiclav Cost Per Prescription Item",
    subtitle = "Monthly average cost per item (Feb 2021 – Jan 2026)",
    x = "Date",
    y = "Cost per item"
  ) +
  theme_minimal()

ggplotly(p_cost_per_item)

# Summary table
cost_per_item_summary <- cost_per_item |> 
  mutate(
    year = year(date),
    month = month(date)
  ) |> 
  group_by(year, month) |> 
  summarise(
    avg_cost_per_item = mean(cost_per_item, na.rm = TRUE),
    min_cost = min(cost_per_item, na.rm = TRUE),
    max_cost = max(cost_per_item, na.rm = TRUE),
    .groups = "drop"
  )

kable(cost_per_item_summary, digits = 2, caption = "Average cost per prescription item by year")
```

## Interpretation: What Caused the Cost Spike?

The dramatic increase in co-amoxiclav costs in late 2022–2023 is not a statistical anomaly but a real-world event with known causes.

### The Strep A Outbreak

Between January 2022 and January 2023, the cost of co-amoxiclav 500mg/125mg tablets surged by **over 500%**. Prices rose from approximately **£1.35–£1.70 per pack to over £8.00–£18.00** in some cases.

The cause was "unprecedented" demand during a **Group A Streptococcus (Strep A) outbreak**, which led to a rapid increase in antibiotic prescriptions. This was compounded by global supply chain shortages for raw materials.

### Impact on the NHS

The surge meant pharmacists were paying significantly more for the medicine than the NHS reimbursement price. The Department of Health and Social Care (DHSC) issued supply warnings, noting that while stock was available with manufacturers, it was not reaching pharmacies fast enough.

### Market Dynamics

This spike fits a recurring pattern: low, often unprofitable, market prices for generics lead to manufacturer withdrawal, followed by shortages and high prices.

### Temporal Alignment with UKHSA Data

The cost spike coincides precisely with a period of elevated Group A Streptococcal (GAS) infections in England. A UK Health Security Agency (UKHSA) report on the 2022/23 season noted a "period of considerable elevation in scarlet fever notifications" and "unusual seasonal patterns" following the COVID-19 pandemic \[1\].

Specifically, UKHSA surveillance data shows that laboratory notifications of invasive GAS (iGAS) infection began to rise sharply in late 2022. The report for the 2025/26 season retrospectively notes that the early part of the 2022/23 GAS season saw a significant upsurge in cases, a trend which is not being observed in the current season \[1\].

This established temporal alignment—between a sharp rise in GAS infections, the subsequent increase in antibiotic demand, and the documented supply shortages—provides the causal mechanism for the 500% price surge.

### Ongoing Volatility

As of 2025, price concessions have continued to be necessary for certain co-amoxiclav products, suggesting ongoing cost volatility.

### Stewardship Implications

| Finding | Implication |
|----|----|
| Prescribing volume stable | The spike was not due to overprescribing |
| Cost spike driven by external factors | Pharmacoeconomic surveillance needed \[6\] |
| Ongoing volatility | Price monitoring should be part of stewardship \[3\] |

**Key takeaway:** The NHS needs systems to monitor not just prescribing volume but also **cost volatility** and **supply chain resilience**.

## Regional Variation

### Top Prescribing Organisations

```{r}
#| label: prescribing-organisations

# Top 10 organisations by total items
top_orgs <- coamox_data |> 
  group_by(organisation_name) |> 
  summarise(
    total_prescribing_items = sum(prescribing_items, na.rm = TRUE),
    .groups = "drop"
  ) |> 
  arrange(desc(total_prescribing_items)) |> 
  head(10)

kable(top_orgs, caption = "Top 10 prescribing organisations")

```

### Key Observations

-   **London dominates prescribing**: 4 of the top 5 prescribing organisations are in London (North East, North West, South East, South West London)
-   **Total prescriptions**: The top 10 organisations account for approximately `r sum(top_orgs$total_prescribing_items)` items
-   **Regional variation**: Significant variation exists, with London ICBs consistently higher than other regions

**Note:** This reflects prescribing volume, not population-adjusted rates. Without population denominators, we cannot determine if higher prescribing is due to larger populations or higher prescribing intensity.

### Identifying East Midlands Organisations

```{r}
#| label: east-midlands-organisations

# List all unique organisation names
all_orgs <- coamox_data |> 
  distinct(organisation_code, organisation_name) |> 
  arrange(organisation_name)

# Search for East Midlands keywords
east_midlands_keywords <- c("NOTTINGHAM", "DERBY", "LEICESTER", "LINCOLN", 
                             "NORTHAMPTON", "RUTLAND", "EAST MIDLANDS")

east_midlands_orgs <- coamox_data |> 
  distinct(organisation_code, organisation_name) |> 
  filter(str_detect(toupper(organisation_name), paste(east_midlands_keywords, collapse = "|")))

# View results
kable(east_midlands_orgs, caption = "East Midlands Organisations")

# East Midlands ICB codes
east_midlands_codes <- east_midlands_orgs$organisation_code

# Calculate East Midlands prescribing over time
east_midlands_data <- coamox_data |> 
  filter(organisation_code %in% east_midlands_codes) |> 
  group_by(date) |> 
  summarise(
    em_prescribing_items = sum(prescribing_items, na.rm = TRUE),
    em_cost = sum(prescribing_cost, na.rm = TRUE),
    .groups = "drop"
  )

# Join with national data
regional_comparison <- national_trend |> 
  left_join(
    east_midlands_data,
    by = "date"
    ) |> 
  mutate(
    em_share_prescribing_items = em_prescribing_items / total_prescribing_items * 100,
    em_share_cost = em_cost / total_cost * 100
  )

# Average share
avg_em_share <- mean(regional_comparison$em_share_prescribing_items, na.rm = TRUE)
expected_share <- 100/9 # 11.11%

# Plot
ggplot(regional_comparison, aes(x = date, y = em_share_prescribing_items)) +
  geom_line(color = "darkred", linewidth = 1) +
  geom_hline(yintercept = expected_share, linetype = "dashed", color = "gray50") +
  annotate("text", x = min(regional_comparison$date), y = expected_share + 1,
           label = paste0("Expected share (", round(expected_share, 1), "%)"),
           hjust = 0, size = 3) +
  labs(
    title = "East Midlands Share of Co-amoxiclav Prescribing",
    subtitle = paste0("Average share: ", round(avg_em_share, 1), "% vs expected 11.1%"),
    x = "Date",
    y = "Percentage of national prescriptions"
  ) +
  theme_minimal()

  
```

### Link to C. difficile Outbreak

Previous analysis detected a **sustained C. difficile outbreak signal in the East Midlands**, with 80% of weeks exceeding the Poisson threshold during 2024.

However, co-amoxiclav prescribing in the East Midlands accounts for only **`r round(avg_em_share, 1)`%** of national prescribing – consistently **below** the expected share of 11.1%.

This suggests that:

1.  **Co-amoxiclav use alone does not explain the outbreak signal**
2.  Other factors should be investigated:
    -   **Antibiotic course length** (not just volume)
    -   **Other high-risk antibiotics** (cephalosporins, quinolones)
    -   **Infection control practices**
    -   **Patient case mix**

**Stewardship implication:** While co-amoxiclav prescribing is not elevated, the East Midlands outbreak warrants broader investigation of antibiotic use patterns and infection control measures.

## References

1.  UK Health Security Agency. (2026). *Group A streptococcal infections: report on seasonal activity in England, 2025 to 2026*. GOV.UK. \[[link](https://www.gov.uk/government/publications/group-a-streptococcal-infections-seasonal-activity-in-england-2025-to-2026/group-a-streptococcal-infections-report-on-seasonal-activity-in-england-2025-to-2026)\]

2.  Chemist+Druggist. (2023). *Co-amoxiclav price surges amid Strep A outbreak*.

3.  Department of Health and Social Care. (2022-2023). *Supply notifications for co-amoxiclav*.

4.  British Generic Manufacturers Association. (2023). *Generic medicine supply challenges*.

5.  OpenPrescribing.net, Bennett Institute for Applied Data Science, University of Oxford.

6.  NHS England. (2024). *National Medicines Optimisation Opportunities 2024/25*.
